---
title: "[Advent of Code 2020 - Day 4](https://adventofcode.com/2020/day/4)"
author: "parmsam"
date: "12/22/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)
library(stringr)

setwd("/Users/samparmar/Documents/programming/advent of code/aoc-2020")
```

## Part A
### Question: Count the number of valid passports - those that have all required fields. Treat cid as optional. In your batch file, how many passports are valid?
```{r}
a <- readr::read_file("day4input.txt")
b <- str_split(a,"\\n\\n")
c <- data.frame(b) %>% rename("from" = 1)

#helpful example: https://stackoverflow.com/questions/33288695/how-to-use-tidyrseparate-when-the-number-of-needed-variables-is-unknown
d <- c %>% 
  mutate(passport_numb = row_number()) %>% 
  mutate(value = strsplit(from, regex(" |\\n"))) 

e <- d %>% unnest(value) 

f <- e %>% 
  group_by(passport_numb) %>% 
  filter(!str_detect(value,"cid")) %>% 
  count() %>% 
  filter(n>6)

f %>% ungroup() %>% select(passport_numb) %>% count()
```
### Answer: 254 valid passports in the batch file for part A

## Part B
### Question: Count the number of valid passports - those that have all required fields and valid values. Continue to treat cid as optional. In your batch file, how many passports are valid?
```{r}
f2 <- e %>% 
  separate(value, c("field","values"), sep=":") %>% 
  ungroup() %>% 
  filter(!is.na(field)) %>% 
  spread(field, values)

f2 <- f2 %>% 
  mutate(byr = as.numeric(byr), 
         cid = as.numeric(cid), 
         eyr = as.numeric(eyr), 
         iyr = as.numeric(iyr))


#cid (Country ID) - ignored, missing or not.

f2 <- f2 %>% 
  select(passport_numb, byr, iyr, eyr, hgt, hcl, ecl, pid) %>% 
  mutate(hgt_unit=str_extract(hgt, "[A-Za-z]+")) %>% 
  mutate(hgt=as.numeric(str_extract(hgt, "[0-9]+"))) %>% 
  select(passport_numb, byr, iyr, eyr, hgt, hgt_unit, everything())

# byr (Birth Year) - four digits; at least 1920 and at most 2002. - DONE
# iyr (Issue Year) - four digits; at least 2010 and at most 2020. - DONE
# eyr (Expiration Year) - four digits; at least 2020 and at most 2030. - DONE
# hgt (Height) - a number followed by either cm or in:
# If cm, the number must be at least 150 and at most 193. *
# If in, the number must be at least 59 and at most 76.   *
# hcl (Hair Color) - a # followed by exactly six characters 0-9 or a-f. - DONE
# ecl (Eye Color) - exactly one of: amb blu brn gry grn hzl oth. - DONE
# pid (Passport ID) - a nine-digit number, including leading zeroes. - DONE

f3 <- f2 %>% filter(byr>=1920, byr<=2002, 
              iyr>=2010, iyr<=2020, 
              eyr>=2020, eyr<=2030,
              str_detect(hgt_unit, "cm|in"),
              (str_detect(hgt_unit, "cm") & (hgt>=150 & hgt<=193)) | (str_detect(hgt_unit, "in") & (hgt>=59 & hgt<=76)),
              str_detect(hcl,"#[0-9a-f]{6}\\b"), 
              ecl %in% c("amb","blu","brn","gry","grn","hzl","oth"), 
              str_detect(pid,"\\b[0-9]{9}\\b"))

f3 %>% count()
```
### Answer: 184 valid passports in the batch file for part B



